---
title: "Lab 4 : Advanced ggplot"
author: "Jeremy Morris"
date: "2/21/2019"
output: 
  html_document: 
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(tidyverse)
library(here)
library(nlme)
```


# Components of a basic plot

# create initial plot
```{r basic}
ggplot(data = mpg, aes(x=displ, y=hwy, colour = factor(cyl))) +
  geom_point()
```

What is happening with above call to ggplot? The first part of the call maps aesthetics to data. The call to `summary` helps us look at the plot object.

```{r basic_summary}
p <- ggplot(data = mpg, aes(x=displ, y=hwy, colour = factor(cyl)))
summary(p) 
```

The resulting mapping is as follows (showing the top 10 rows only)
```{r mapping}
head(data.frame(x=mpg$displ, y=mpg$hwy, colour=factor(mpg$cyl)), n = 10)
```

Note that it can't be plotted yet because there are no layers at this point
```{r basic_plot}
p 
```

To add a layer we add a geometric object (a.k.a geom); in this case it's a point geom

```{r add_geom}
p <- p + geom_point()
summary(p) # examine the components of the plot
```

Now that there is a layer, it can be plotted. Note that we can save a plot to a variable and then plot it by calling just that variable. We can also assign a base visualization to a variable and then add to it later. We'll see that a little as we go on.
```{r plot}
p # plot it
```

Applying other geoms; the geom type determines the type of the plot
line: grammatically correct, but doesn't make sense

```{r line_wrong}
ggplot(data = mpg, aes(x=displ, y=hwy, colour = factor(cyl))) +
  geom_line()
```

bar: grammatically correct, but doesn't make sense

```{r bar_wrong}
ggplot(data = mpg, aes(x=displ, y=hwy, colour = factor(cyl))) +
  geom_bar(stat="identity") 
```

Refer to lab 2 for a list of all `geoms`

Lets go back to the original scatter plot and add some regressions
```{r add_regression1}
p <- ggplot(data = mpg, aes(x=displ, y=hwy, colour = factor(cyl))) +
  geom_point()
```

Now create an additional layer with a regression. Note how we're adding `geom_smooth` to the existing plot.
```{r add_regression2}
p + geom_smooth()
```

What happens if we add the mappings at the geom level?
```{r add_to_geom}
ggplot(data = mpg) +
  geom_point(aes(x=displ, y=hwy, colour = factor(cyl))) +
  geom_smooth(aes(x=displ,y=hwy))
```

# scaling and coordinates
# We need to convert from data units (e.g., litres, miles per gallon and number 
# of cylinders) to physical units (e.g., pixels and colours) that the computer 
# can display. This conversion process is called scaling and performed by scales.

# linear data is mapped to a [0,1] scale

# coordinate system in ggplot2 maps to pixels
# consider: cartesian coordinates, polar coordinates, color space

# To create a complete plot we need to combine graphical objects from three 
# sources: the data, represented by the point geom; the scales and coordinate 
# system, which generate axes and legends so that we can read values from the 
# graph; and plot annotations, such as the background and plot title.

# below represents the scaling from [0,1] for the continuous variables 
# (displ, hwy) and the colour mapping for the variable factor(cyl)

# x     y     colour  size shape
# ￼0.037 0.531 #FF6C91 1 19
# 0.037 0.531 #FF6C91 1 19 
# 0.074 0.594 #FF6C91 1 19 
# 0.074 0.562 #FF6C91 1 19 
# 0.222 0.438 #00C1A9 1 19 
# 0.222 0.438 #00C1A9 1 19 
# 0.278 0.469 #00C1A9 1 19 
# 0.037 0.438 #FF6C91 1 19 
# 0.037 0.406 #FF6C91 1 19 
# 0.074 0.500 #FF6C91 1 19

# a final look at the plot
ggplot(data = mpg, aes(x=displ, y=hwy, colour = factor(cyl))) +
  geom_point()

# components of a more advanced plot including multiple layers, facets, and stats ####
# start with a new plot
ggplot(data = mpg, aes(x=displ, y=hwy)) +
  geom_point() +
  geom_smooth()

# note that stat_smooth is an alias for geom_smooth
ggplot(data = mpg, aes(x=displ, y=hwy)) +
  geom_point() +
  stat_smooth() # this is the same plot

# now create a facet by year
p <- ggplot(data = mpg, aes(x=displ, y=hwy,color=factor(cyl))) + 
  geom_point() +
  # geom_smooth() +
  facet_grid(. ~ year)
p # plot it

ggplot(data = mpg, aes(x=displ, y=hwy)) + 
    geom_point() +
    # geom_smooth() +
    facet_grid(cyl ~ year,scales='free_y')

# now change the grid to be vertical
p <- ggplot(data = mpg, aes(x=displ, y=hwy)) + 
  geom_point() +
  geom_smooth() +
  facet_grid(year ~ .)
p # plot it

summary(p) # examine the components of the plot

# now change the default method to lm
p <- ggplot(data = mpg, aes(x=displ, y=hwy)) + 
  geom_point() +
  geom_smooth(method = "lm") +
  facet_grid(year ~ .)
p # plot it

summary(p) # examine the components of the plot

# review related slides - second time ####
# see: AUX_lab-04_advanced-ggplot2_v1.x.pptx

# build a plot layer by layer ####
p <- ggplot(diamonds, aes(x=carat))
summary(p)
p # recall that we cannot render it yet without a layer

# add a layer the long way
p <- p + layer(
  geom = "bar"
  , geom_params = list(fill = "steelblue")
  , stat = "bin"
  , stat_params = list(binwidth = 2)
)
summary(p)
p # now we can render it

# add a layer the the short way
p <- ggplot(diamonds, aes(x=carat)) +
  geom_histogram(binwidth = 2, fill = "steelblue")
summary(p)
p # rendering is the same

# geom_histogram() have default values for their geom and are therefore
# easier/faster to use as compared to the layer() call; geom_XXX are called
# shortcut functions

# overriding default mappings in layer ####
p <- ggplot(mtcars, aes(x = mpg, y=wt))
p <- p + geom_point()
p

# override default color
p <- p + geom_point(aes(colour=factor(cyl)))
p # this makes sense

# override x mapping
p <- p + geom_point(aes(x=disp))
p # looks strange; notice the extra points and the larger range of x

# setting an aesthetic vs. mapping it ####
p <- ggplot(mtcars, aes(x=mpg, y=wt))
p <- p + geom_point(colour = "darkblue") # example of setting (vs. mapping)
summary(p)
p

# now try mapping but using the constant value of "darkblue"
p <- ggplot(mtcars, aes(x=mpg, y=wt))
p <- p + geom_point(aes(colour = "darkblue")) # example of mapping (vs. setting)
summary(p)
p # why isn't it dark blue?

# "This effectively creates a new variable containing only the value “darkblue” 
# and then maps colour to that new variable. Because this value is discrete, the 
# default colour scale uses evenly spaced colours on the colour wheel, and since 
# there is only one value this colour is pinkish."

# grouping observations ####
# geoms can be roughly divided into individual and collective geoms. An individual 
# geom has a distinctive graphical object for each row in the data frame. For example, 
# the point geom has a single point for each observation. On the other hand, 
# collective geoms represent multiple observations. This may be a result of a 
# statistical summary, or may be fundamental to the display of the geom, as 
# with polygons. Lines and paths fall somewhere in between: each overall line 
# is composed of a set of straight segments, but each segment represents two points.


# examine the data 
?Oxboys
summary(Oxboys)
str(Oxboys)

# plot a line per subject
p <- ggplot(Oxboys, aes(x=age, y=height)) +
  geom_line()
p # this looks strange; the line is going through all of the points instead
# or rendering a line per subject

# correct this using the group aesthetics
p <- ggplot(Oxboys, aes(x=age, y=height, group=Subject)) +
  geom_line()
p
summary(p)

# now plot a linear fit for the whole dataset
p <- ggplot(Oxboys, aes(age, height, group = Subject)) +
  geom_line() + 
  geom_smooth(method="lm", se=FALSE)
p

# use group = 1 to ungroup from the main call to ggplot
p <- ggplot(Oxboys, aes(age, height, group = Subject)) +
  geom_line() + 
  geom_smooth(aes(group = 1), method="lm", se=FALSE)
p

  
